\documentclass{jfp1}

\usepackage{color}

\title{The GHC Runtime System}
\author{fill in}

\newcommand{\Red}[1]{{\color{red} #1}}
\newcommand{\ToDo}[1]{{\color{blue} ToDo: #1}}
\newcommand{\SM}[1]{{\color{green} SM: #1}}
\newcommand{\nb}[1]{$\lhd$ \Red{#1} $\rhd$}
\newcommand{\tocite}[0]{{\color{red} [cite]}\xspace}

\newcommand{\XXX}{\Red{XXX}}

\begin{document}

\maketitle

\begin{abstract}
(This paper describes the implementation of the GHC runtime system)
\end{abstract}

\tableofcontents

\section{Introduction}

The GHC runtime system is perhaps one of the most unusual language
runtimes in wide use today.  Much of its implementation was directly
motivated by the unusual (by popular programming language standards!)
features in Haskell which the runtime needs to support.  Some of these
features, such as lazy evaluation and ubiquitous concurrency, complicate
the design of the system and require us to do a lot more work to give
programmers the high-level behavior they desire.  Other features, such
as an emphasis on pure computation without mutation, simplify the design
components such as garbage collection and software transactional memory.

\Red{Complete, flesh out, rewrite}

\SM{This is a good start.  I think it will help to tell the reader up
  front that we intend to focus on the main novel aspects of the
  runtime how concurrency and parallelism, the garbage collector, and
  lazy evaluation work.  Everything else we say will be stuff that you
  need to know in order to understand these three main topics, and the
  order of the paper will reflect this - talking about the essential
  aspects of the evaluation model and data representations first,
  because you need that to understand the rest.}

\SM{Please feel free to label sections that you would like me to
  write.}

\input{evaluation.tex}
\input{storage.tex}
\input{threads.tex}

\section{Acknowledgements}

Alexander Chernyakhovsky for helpful discussions.

\bibliographystyle{jfp}
\bibliography{paper}

\end{document}
